---
title: "Comparison of Ebola virus glycoprotein region between Ebola Zaire and Ebola Reston"
output: html_notebook
---

[here we can put the theory behind our project and code]

```{r}
# Loading libraries

library(Biostrings)
library(tidyverse)
library(pwalign)
library(ggthemes)
library(ggplot2)
library(dplyr)
library(ggmsa)
library(ggseqlogo)

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

```{r}
# Sequence alignment 

# Using a stretch of amino acid sequences from AA 30 to 60, spanning AA 33-50 
# AA 33-50 has been identified as playing a role in infectivity 

# Read in sequences from txt files
# Zaire GP
zgp <- AAString(readLines("zgp_30_60.txt"))
print(zgp)

# Reston GP 
rgp <- AAString(readLines("rgp_30_60.txt"))
print(rgp)

 
# Align sequences using pairwise alignment 
alignment_GP <- pairwiseAlignment(pattern = zgp, subject = rgp)
print(alignment_GP)


# Extract sequences (with ziare as pattern and reston as subject) and convert to character vector
# Character vector will store strings of text 
aligned_pattern <- as.character(alignment_GP@pattern)
aligned_subject <- as.character(alignment_GP@subject)

# Create alignment matrix 
# Splits into single characters
# 2 rows, one for each sequence
# byrow = TRUE --> data filled by row 
alignment_matrix <- matrix(c(strsplit(aligned_pattern, "")[[1]],
                             strsplit(aligned_subject, "")[[1]]),
                           nrow = 2, byrow = TRUE)


# Labels rows, or sequences, as below 
rownames(alignment_matrix) <- c("Zaire", "Reston")
```


When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

