---
title: "Comparative proteomics of lethal and non-pathogenic Ebolavirus species envelope glycoproteins"
output: html_notebook
---

**A Critical Domain of Ebolavirus Envelope Glycoprotein Determines Glycoform and Infectivity** by Fujihira et al, 2018 (https://www.nature.com/articles/s41598-018-23357-8) identifies a small stretch of surface glycoprotein amino acids (amino acids 33-50) that play a significant role in the glycoprotein's ability to infect cells in vitro. They found differences in this stretch between Zaire ebolavirus and Reston ebolavirus. Using an in vitro system and the glycoprotein's known receptor, they found that Reston ebolavirus became more infectious when this stretch of it's glycoprotein was replaced with that of Zaire ebolavirus.

We chose to recapitulate this seqeuence alignment and analyze the amino acid differences.

```{r}
# Loading libraries

library(Biostrings)
library(tidyverse)
library(pwalign)
library(ggthemes)
library(ggplot2)
library(dplyr)
library(ggmsa)
library(ggseqlogo)

```

## Sequence alignment 
Using a stretch of amino acid (AA) sequences from AA 30 to 60, spanning AA 33-50 
AA 33-50 has been identified as playing a role in infectivity 

```{r}
# Read in sequences from txt files
# Zaire GP
zgp <- AAString(readLines(file.choose())) # Choose zgp_30_60.txt
print(zgp)

# Reston GP 
rgp <- AAString(readLines(file.choose())) # Choose rgp_30_60.txt
print(rgp)

 
# Align sequences using pairwise alignment 
alignment_GP <- pairwiseAlignment(pattern = zgp, subject = rgp)
print(alignment_GP)


# Extract sequences (with ziare as pattern and reston as subject) and convert to character vector
# Character vector will store strings of text 
aligned_pattern <- as.character(alignment_GP@pattern)
aligned_subject <- as.character(alignment_GP@subject)

# Create alignment matrix 
# Splits into single characters
# 2 rows, one for each sequence
# byrow = TRUE --> data filled by row 
alignment_matrix <- matrix(c(strsplit(aligned_pattern, "")[[1]],
                             strsplit(aligned_subject, "")[[1]]),
                           nrow = 2, byrow = TRUE)


# Labels rows, or sequences, as below 
rownames(alignment_matrix) <- c("Zaire", "Reston")

```


## ggmsa visualization

```{r}
# Convert matrix to data frame 
alignment_aa <- AAStringSet(apply(alignment_matrix, 1, paste, collapse = ""))

# Ensure unique names for each sequence 
names(alignment_aa) <- rownames(alignment_matrix)

# Plot

ggmsa <- ggmsa(alignment_aa) +
  theme_minimal() + 
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 14)) +
  labs(title = "Amino acids 30-60 comparison - Ebola Zaire and Ebola Reston GP",  # Title of the plot
    x = "Position",  # X-axis label
    y = "Amino Acid"  # Y-axis label
  )
plot(ggmsa)
#ggsave("ggmsa_GP_alignment.png", plot = ggmsa, width = 10, height = 6, dpi = 300)

```


## ggseqlogo visualization 

```{r}
# Position frequency matrix, shows how often the sequence appears in that position of the alignment
create_pfm <- function(mat) {
  unique_aa <- unique(unlist(mat)) # Finds unique AAs in vector form
  pfm <- matrix(0, nrow = length(unique_aa), ncol = ncol(mat)) # Makes empty matrix with 0s for now 
  rownames(pfm) <- unique_aa # Assigns the unique AAs to the two rows 
  
  # Loop through each column in the sequence alignment matrix 
  for (i in 1:ncol(mat)) {
    amino_acids <- mat[, i]
    for (aa in unique_aa) { # Iterates over each AA to see if it's present in both rows of the column 
      if (aa %in% amino_acids) {
        pfm[aa, i] <- sum(amino_acids == aa) # Sums probability/frequency of those AAs in the column
      }
    }
  }
  
  return(pfm) # Prints resulting matrix 
}

# Creates data frame for the position frequency matrix 
pfm <- create_pfm(alignment_matrix)


# Create sequence logo using pfm 

ggseqlogo <- ggseqlogo(pfm, method = "prob") + 
  labs(x = "Position", y = "Probability", 
       title = "Amino acids 30-60 comparison - Ebola Zaire and Ebola Reston GP") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 12), 
        axis.text = element_text(size = 10))
plot(ggseqlogo)
#ggsave("ggseqlogo_GP_alignment.png", plot = ggseqlogo, width = 10, height = 1.5, dpi = 300)

```


## Create a substitution table and visualize frequency amino acid changes

```{r}
# Create data frame for table
substitution_table <- data.frame(From = character(0),
                                 To = character(0),
                                 Frequency = integer(0),
                                 stringsAsFactors = FALSE)

# Loop through alignment to compare amino acids and see if any have changed 
for (i in 1:nchar(aligned_pattern)) {  
  aa1 <- substr(aligned_pattern, i, i)  # Amino acid from Zaire sequence (pattern)
  aa2 <- substr(aligned_subject, i, i)  # Amino acid from Reston sequence (subject)
  
  if (aa1 != aa2) {  # Tracks changes, not matches, where aa1 does not equal aa2
    # Add the substitution to the table (From -> To)
    substitution_table <- rbind(substitution_table, # Fills substitution table 
                                data.frame(From = aa1, 
                                           To = aa2, 
                                           Frequency = 1))
  }
}

# Calculates frequency of each substitution and puts into table format
substitution_table <- substitution_table %>%
  group_by(From, To) %>%
  summarise(Frequency = sum(Frequency)) %>%
  arrange(desc(Frequency))
#view(substitution_table)


# Plot substitution table 
sub_table <- ggplot(substitution_table, aes(x = interaction(From, To), y = Frequency, fill = Frequency)) +
  geom_bar(stat = "identity", show.legend = FALSE, width = 0.6) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  scale_y_continuous(breaks = seq(0, max(substitution_table$Frequency), by = 1)) +
  theme_minimal() +
  labs(x = "Substitution (From → To)", 
       y = "Frequency", 
       title = "Substitution Frequencies Between Zaire and Reston GP") +
  theme(axis.text.x = element_text(hjust = 1),
        panel.grid.major = element_line(color = "gray", size = 0.5),
        panel.grid.minor = element_blank()  
  )

plot(sub_table)
ggsave("substitution_table.png", plot = sub_table, width = 10, height = 5, dpi = 300)
```



## Find alignment percent identity 

```{r}
# Assigns matches as 0 
matches <- 0
total_positions <- nchar(aligned_pattern)

# Loops through each position in alignment and checks for matches 
for (i in 1:total_positions) {
  aa1 <- substr(aligned_pattern, i, i)  # Zaire sequence
  aa2 <- substr(aligned_subject, i, i)  # Reston sequence
  
  if (aa1 == aa2) {  # Check if the amino acids are the same
    matches <- matches + 1 # If match, add to matches 
  }
}

# Percent identity calculation 
percent_identity <- (matches / total_positions) * 100

# Print
cat("Percent Identity between Zaire and Reston GP: ", percent_identity, "%\n")
```


## Looking at amino acid properties
This will look at single changes at a time, meaning we can go from our substitution table and ask this block to look at specific changes and the nature of them. 
For example, if we want to investigate V --> I, which shows up twice above, we can do this below. 


```{r}
# Define properties for each amino acid
# This was taken from a list provided by ChatGPT, not generated by us 
aa_properties <- data.frame(
  Amino_Acid = c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L", "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y"),
  Charge = c("Neutral", "Neutral", "Acidic", "Acidic", "Neutral", "Neutral", "Basic", "Neutral", "Basic", "Neutral", "Neutral", "Polar", "Neutral", "Polar", "Basic", "Neutral", "Neutral", "Neutral", "Neutral", "Neutral"),
  Size = c("Small", "Small", "Medium", "Medium", "Large", "Small", "Medium", "Large", "Large", "Large", "Medium", "Small", "Small", "Small", "Large", "Small", "Small", "Small", "Large", "Large"),
  Hydrophobicity = c("Hydrophobic", "Hydrophobic", "Hydrophilic", "Hydrophilic", "Hydrophobic", "Hydrophobic", "Hydrophilic", "Hydrophobic", "Hydrophilic", "Hydrophobic", "Hydrophobic", "Hydrophilic", "Hydrophobic", "Hydrophilic", "Hydrophilic", "Hydrophilic", "Hydrophilic", "Hydrophobic", "Hydrophobic", "Hydrophobic")
)

# Most frequent amino acid substitution (V --> I)
from_aa <- "V"
to_aa <- "I"

# Get properties for the substitution
from_properties <- aa_properties[aa_properties$Amino_Acid == from_aa, ]
to_properties <- aa_properties[aa_properties$Amino_Acid == to_aa, ]

# Print the properties of both amino acids
cat("From:", from_aa, "Properties:", paste(from_properties[2:4], collapse = ", "), "\n")
cat("To:", to_aa, "Properties:", paste(to_properties[2:4], collapse = ", "), "\n")

```





